Type-Safe Modular Parsing

### 2017-01-13 paper/sections/CaseStudy.tex
Fix sec 4 5 6

\paragraph{Dependency Overview}
Figure \ref{fig:dependency} shows the
dependency of all the components and calculi in our case study. Gray
boxes are calculi and white boxes are components. An arrow starting
from box A to box B denotes that B includes and thus reuses A.

@@

Table \ref{tab:ext-comparison} shows part of the comparison results of these four versions,
full table is available online []. It suggests that the difference of running time between
using object algebra and class is little, roughly 1\%.
The usage of longest match combinator slows the performance by 7\%. The main reason of slower
execution time may be the overall structure of the modular parsing approach, because we indeed have
more intermediate function calls and method overriding. However, it is worth metioning that
because of the memoization technique of Packrat parsers, we are only constant times
slower, the algorithmic complexity is still the same.

>>>>

\paragraph{Dependency Overview}
Figure \ref{fig:dependency} shows the
dependency of all the components and calculi in our case study. Grey
boxes are calculi and white boxes are components. An arrow starting
from box A to box B denotes that B includes and thus reuses A.

@@

Table \ref{tab:ext-comparison} shows part of the comparison results of these four versions,
full table is available online []. It suggests that the difference of running time between
using object algebra and class is little, roughly 1\%.
The usage of longest match combinator slows the performance by 7\%. The main reason of slower
execution time may be the overall structure of the modular parsing approach, because we indeed have
more intermediate function calls and method overriding. However, it is worth mentioning that
because of the memoization technique of Packrat parsers, we are only constant times
slower, the algorithmic complexity is still the same.

### 2017-01-13 paper/sections/Discussion.tex
Fix sec 4 5 6

Another advantage of using Object Algebras is that is is easy to
model multi-sorted languages. If the syntax contains multiple sorts, we can distinguish them by different type parameters. As an example, we extend the language of literals, additions and variables by introducing types as a new kind of type argument, and we add lambda abstractions to expressions. Now the language has two sorts in the syntax, types and expressions:\\

@@

From the code we observe that the multi-sorts of Object Algebra interface
express multiple syntax. The isolation for different syntax
is guaranteed by the type system. The following illustrates some client code
for the new parser:

@@

One can even ``eliminate'' a production rule in the extension, by actually overriding it with a failure parser. The lexer can also be updated, since keywords and delimiters are represented by sets of strings.

@@

As shown below, those two language components can be merged into one to obtain the language we want. Futhermore, the new language is still a modular
component ready for future composition. In that case modularity is realized over higher-order hierarchies.

\lstinputlisting[linerange=160-170]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR
Such composition shows perfect modularity. The only
drawback is that the glue code of composition appears to be boilerplate. As shown above, we are combining ASTs, parsers and pretty-printers of
\lstinline{VarExpr} and \lstinline{TypedLam} respectively. Such a pattern refers to \textit{family polymorphism}[]. Unfortunately Scala does not
support this feature, nonetheless, one can avoid such boilerplate by using metaprogramming techniques.

@@

We use traits to model parsers, OO inheritance to compose them, and method overriding for their extensibility. These language features may not be supported in other programming languages, especially in functional languages such as Haskell. Nevertheless, \textit{open recursion} could be used as an alternative. An explict ``self-reference'' parameter is able to explain the recursive calls dynamically, by the real argument passed at runtime.

@@

Besides Object Algebras, many other techniques including \textit{Data types à la carte} (DTC) [] and \textit{Modular reifiable matching} (MRM) [] also support extensible data structures. These alternatives can also be adopted to build ASTs and corresponding parsers.

>>>>

Another advantage of using Object Algebras is that it is easy to
model multi-sorted languages. If the syntax contains multiple sorts, we can distinguish them by different type parameters. As an example, we extend the language of literals, additions and variables by introducing types as a new kind of type argument, and we add lambda abstractions to expressions. Now the language has two sorts in the syntax, types and expressions:\\

@@

From the code, we observe that the multi-sorts of Object Algebra interface
express multiple syntax. The isolation for different syntax
is guaranteed by the type system. The following illustrates some client code
for the new parser:

@@

One can even ``eliminate'' a production rule in the extension, by overriding it with a failure parser. The lexer can also be updated, since keywords and delimiters are represented by sets of strings.

@@

As shown below, those two language components can be merged into one to obtain the language we want. Furthermore, the new language is still a modular
component ready for future composition. In that case modularity is realized over higher-order hierarchies.

\lstinputlisting[linerange=160-170]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR
Such composition shows perfect modularity. The only
drawback is that the glue code of composition appears to be boilerplate. As shown above, we are combining ASTs, parsers and pretty-printers of
\lstinline{VarExpr} and \lstinline{TypedLam} respectively. Such a pattern refers to \textit{family polymorphism}~\cite{ernst01FP} which is unfortunately not supported in Scala. Nonetheless, one can avoid such boilerplate by using metaprogramming techniques.

@@

We use traits to model parsers, OO inheritance to compose them, and method overriding for their extensibility. These language features may not be supported in other programming languages, especially in functional languages such as Haskell. Nevertheless, \textit{open recursion} could be used as an alternative. An explicit ``self-reference'' parameter is able to explain the recursive calls dynamically, by the real argument passed at runtime.

@@

Besides Object Algebras, many other techniques including \textit{Data types à la carte} (DTC)~\cite{swierstra2008data} and \textit{Modular reifiable matching} (MRM)~\cite{Oliveira2015} also support extensible data structures. These alternatives can also be adopted to build ASTs and corresponding parsers.

### 2017-01-13 paper/sections/ObjectAlgebras.tex
Fix sec 4 5 6

The inheritance-based approach allows building extensible parsers, based on an OO class hierarchy.  Nevertheless, the addition of new operations over ASTs is problematic using traditional OO ASTs. In this section, we
show how to support both the forms of extensibility for ASTs (easy
addition of language contructs, and easy addition of operations)
using Object Algebras~\cite{Oliveira2012}.

@@

The Expression Problem illustrates the difficulty of extending data structures or ASTs in two dimensions. For traditional OO ASTs, adding new language constructs is easy, but adding new operations is hard. In the last section, we have a simple language containing literals, additions and variables. The corresponding ASTs are built using classes with a pretty-printing operation. Now suppose we want to collect free variables over the ASTs. By traditional OO approach, one attempt would be extending \inlinecode{Expr} with the new operation to obtain a new abstract type for ASTs:

@@

\subsection{Object Algebras}\label{subsec:objectalgebras}
Object Algebras captures a design pattern to address the Expression Problem, achieving two dimensions of extensibility (language constructs and operations) in a modular and type-safe way. The definition of data structures is separated from their behaviors, and future extensions on both dimensions no longer require existing code to be modified, supporting separate compilation.

Using Object Algebras in Scala, ASTs as recursive data structures are defined by traits, where each constructor corresponds to an abstract method inside.
Essentially, Object Algebras generalize the {\sc Abstract Factory} pattern~\cite{}, and promote the use of factory methods, instead of constructors, for instantiating objects. The example from Section~\ref{subsec:packratparsing} is used here again for illustration.
At first the language only supports literals and additions:

@@

\lstinputlisting[linerange=12-15]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_PRINT
Here \inlinecode{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
One can define evaluation operation as a new trait on this expression langauge:

@@

\paragraph{Adding New AST Constructs}
Futhermore, new language constructs can be added by extending \inlinecode{ExprAlg} and adding new cases only. Now we extend the language
with variables. A new Object Algebra interface \inlinecode{VarAlg} is defined as follows:

@@

Parsing builds the AST as a result. When Object Algebras are used
to build ASTs, an Object Algebra containing the constructor/factory
methods has to be used by the parsing function. Thus, a first attempt
at defining the parser for the small arithmetic language is:

@@

\lstinputlisting[linerange=8-16]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser works fine, but it is not extensible! For example, we have demonstrated in Section~\ref{subsec:overriding} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method override for \inlinecode{pExpr} would require a type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]}, where the type of the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This difference prohibits method/field overriding. Scala supports type-refinement of fields if the overriding
field has a type which is a \emph{subtype} of the original field.
Unfortunatelly, in this case, although \inlinecode{VarExprAlg} is a subtype
of \inlinecode{ExprAlg}, the function type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]} is a supertype of \inlinecode{ExprAlg[E] =>} \inlinecode{Parser[E]}. Thus it is not possible to have overriding.

@@

\subsection{Composition of Algebras}

By adopting Object Algebras, we also inherit some advanced features from the pattern. Specifically, in the original paper~\cite{Oliveira2012}, there are two kinds of Object Algebras: \textit{queries} and \textit{transformations}. A query traverses the abstract syntax tree, collects information and returns the result. The aforementioned \lstinline{Print} and \lstinline{Eval} are traditional queries. But as we discussed before, an algebra is first fed to the parser, then \lstinline{parse} is invoked to do the parsing. It is tedious to fed \lstinline{Print} and \lstinline{Eval} separately, as the same input is parsed twice. Instead, we can firstly compose two queries into one, then feed it to the parser. For space reasons we omit the code here.

>>>>

The inheritance-based approach allows building extensible parsers, based on an OO class hierarchy.  Nevertheless, the addition of new operations over ASTs is problematic using traditional OO ASTs. In this section, we
show how to support both the forms of extensibility for ASTs (easy
addition of language constructs, and easy addition of operations)
using Object Algebras~\cite{Oliveira2012}.

@@

The Expression Problem~\cite{wadler1998expression} illustrates the difficulty of extending data structures or ASTs in two dimensions. For traditional OO ASTs, adding new language constructs is easy, but adding new operations is hard. In the last section, we have a simple language containing literals, additions and variables. The corresponding ASTs are built using classes with a pretty-printing operation. Now suppose we want to collect free variables over the ASTs. By traditional OO approach, one attempt would be extending \inlinecode{Expr} with the new operation to obtain a new abstract type for ASTs:

@@

\subsection{Object Algebras}\label{subsec:objectalgebras}
Object Algebras captures a design pattern to address the Expression Problem, achieving two dimensions of extensibility (language constructs and operations) in a modular and type-safe way. The definition of data structures is separated from their behaviours, and future extensions on both dimensions no longer require existing code to be modified, supporting separate compilation.

Using Object Algebras in Scala, ASTs as recursive data structures are defined by traits, where each constructor corresponds to an abstract method inside.
Essentially, Object Algebras generalize the {\sc Abstract Factory} pattern~\cite{gamma1995design}, and promote the use of factory methods, instead of constructors, for instantiating objects. The example from Section~\ref{subsec:packratparsing} is used here again for illustration.
At first the language only supports literals and additions:

@@

\lstinputlisting[linerange=12-15]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_PRINT
Here \inlinecode{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
One can define evaluation operation as a new trait on this expression language:

@@

\paragraph{Adding New AST Constructs}
Furthermore, new language constructs can be added by extending \inlinecode{ExprAlg} and adding new cases only. Now we extend the language
with variables. A new Object Algebra interface \inlinecode{VarAlg} is defined as follows:

@@

Parsing produces ASTs as the result. When Object Algebras are used
to build ASTs, an Object Algebra containing the constructor/factory
methods has to be used by the parsing function. Thus, a first attempt
at defining the parser for the small arithmetic language is:

@@

\lstinputlisting[linerange=8-16]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser works fine, but it is not extensible! For example, we have demonstrated in Section~\ref{subsec:overriding} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method override for \inlinecode{pExpr} would require a type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]}, where the type of the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This difference prohibits method/field overriding. Scala supports type-refinement of fields if the overriding
field has a type which is a \emph{subtype} of the original field.
Unfortunately, in this case, although \inlinecode{VarExprAlg} is a subtype
of \inlinecode{ExprAlg}, the function type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]} is a supertype of \inlinecode{ExprAlg[E] =>} \inlinecode{Parser[E]}. Thus it is not possible to have overriding.

@@

\subsection{Composition of Algebras}
\huang{to revise}
By adopting Object Algebras, we also inherit some advanced features from the pattern. Specifically, in the original paper~\cite{Oliveira2012}, there are two kinds of Object Algebras: \textit{queries} and \textit{transformations}. A query traverses the abstract syntax tree, collects information and returns the result. The aforementioned \lstinline{Print} and \lstinline{Eval} are traditional queries. But as we discussed before, an algebra is first fed to the parser, then \lstinline{parse} is invoked to do the parsing. It is tedious to fed \lstinline{Print} and \lstinline{Eval} separately, as the same input is parsed twice. Instead, we can firstly compose two queries into one, then feed it to the parser. For space reasons we omit the code here.

### 2017-01-13 paper/sections/RelatedWork.tex
Fix sec 4 5 6

\paragraph{Extensibility} Various design patterns~\cite{}, in multiple
languages, have been proposed over the years to address extensibility
problems, such as the Expression Problem~\cite{}. However, to the best
of our knowledge, none of the existing extensibility techniques
address the problem of \emph{modularly 
parsing extensible ASTs}.  There has been
a lot of work on extensible ASTs, in both functional and
object-oriented programming.  For instance, the famous ``Datatypes \`a
la Carte'' (DTC)
paper~\cite{swierstra2008} represents modular ASTs using co-products
of every two functors, which forms a binary tree structure. 
Several variants of DTC have been later proposed~\cite{Bahr2011,Bahr2014,Oliveira2015}
All of that work essentially covers how to traverse and 
consume extensible ASTs. 
Only in Bahr's~\cite{Bahr2011} work a brief mention of \emph{unfolds} 
(which are a class of algorithms that \emph{build} ASTs) are mentioned. 
However Bahr's work does not cover parsing.

>>>>

\paragraph{Extensibility} Various design patterns~\cite{}, in multiple
languages, have been proposed over the years to address extensibility
problems, such as the Expression Problem~\cite{}. However, to the best
of our knowledge, none of the existing extensibility techniques
address the problem of \emph{modularly
parsing extensible ASTs}.  There has been
a lot of work on extensible ASTs, in both functional and
object-oriented programming.  For instance, the famous ``Datatypes \`a
la Carte'' (DTC)
paper~\cite{swierstra2008data} represents modular ASTs using co-products
of every two functors, which forms a binary tree structure.
Several variants of DTC have been later proposed~\cite{Bahr2011,Bahr2014,Oliveira2015}
All of that work essentially covers how to traverse and
consume extensible ASTs.
Only in Bahr's~\cite{Bahr2011} work a brief mention of \emph{unfolds}
(which are a class of algorithms that \emph{build} ASTs) are mentioned.
However Bahr's work does not cover parsing.
